---
output: pdf_document
---
# 1.0 Executive Summary

For your car to be able to get from point A to point B without having to stay trudging along in first gear, it needs a working transmission. The transmission allows the vehicle to change gears, thereby transferring power from the engine to the drive axle in the most efficient way possible. It does this by varying the gear ratio. In lower gears, this increases available power while reducing speed. Higher gears, on the other hand, reduce power and increase speed. This enables cars to distribute power and speed in the most efficient way for any given situation. But whilst everyone agrees that a transmission is absolutely vital to the inner workings of any car, there is no general consensus regarding what kind of transmission is better-automatic or manual.This document shows outcome of the studies done. 

# 2.0 Data analysis

`mtcars` dataset was used for the analysis. It comprises 10 aspects of automobile design and performance for 32 automobiles.

```{r warning=FALSE}
# factor some variables
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$am <- factor(mtcars$am)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
str(mtcars)
```

The following boxplot shows the relation between the transmission type and the miles per gallon (MPG).

```{r}
library(plyr)
library(ggplot2)
# Rename the levels of transmission types
transmission <- revalue(mtcars$am, c('0'="automatic", '1'="manual"))
ggplot(mtcars, aes(x=transmission, y=mpg, fill=transmission)) +
    geom_boxplot() +
    xlab("Transmission type") +
    ylab("Miles per gallon")
```

It suggests a clear difference on fuel consumption between automatic
and manual transmission cars. Below is the model to explain the MPG
variability with the transmission type _only_.

```{r}
fit1 <- lm(mpg ~ am, data=mtcars)
summary(fit1)
```

Although coefficients for both intercept and the transmission type are
significant, the model fit using _only_ transmission type explains only
`r summary(fit1)$r.squared * 100`% of the MPG variation.



Before making any conclusions on the effect of transmission type on fuel
efficiency, we look at the variances between several variables in the dataset.

```{r, warning=FALSE, fig.width=8, fig.height=6}
pairs(mtcars, panel=function(x,y) {
    points(x, y)
    abline(lm(y ~ x), col="red")
})
```

Based on the pairs plot above, several variables seem to have high
correlation with the `mpg` variable. Hence, we build an initial model
using all variables and select the model with the best subset of
predictors using stepwise backward elimination and forward selection.

```{r}
initial_model <- lm(mpg ~ ., data=mtcars)
best_model <- step(initial_model, direction="both", trace=0)
summary(best_model)
par(mfrow = c(2,2))
plot(best_model)
```

The final model contains four predictors, `cyl` (number of cylinders),
`hp` (horsepower), `weight` (weight) and `am` (transmission type). This
model explains the `r summary(best_model)$r.squared * 100`% of the MPG
variation. The number of cylinders, weight and horsepower  significantly
contribute to the accuracy of the model while the transmission has no effect
on the fuel consumption ($\alpha=0.05$). Also the residual plots show that
the distribution of residuals seem to be
normally distributed and not depending on fitted values.

# 3.0 Results

The data analysis on `mtcars` dataset from 1973 reveals some interesting
points.

- If a car has 6 cylinder or 8 cylinder, rather than 4, the fuel consumption
increases by `r abs(best_model$coefficients[[2]])` and
`r abs(best_model$coefficients[[3]])` MPG, respectively.
- One unit of increase on gross horsepower results
`r abs(best_model$coefficients[[4]])` less MPG, an increase on the fuel
consumption.
- 1000 lb increase on the weight of a car, everything else same, yields
`r abs(best_model$coefficients[[5]])` less MPG, again an increase on the
fuel consumption.

The `mtcars` dataset used for this analysis comprises data for 1973-1974
models. This analysis was not able to find any significant link between the
transmission type and fuel consumption. For modern cars, with much more
efficient automatic transmission system, it is less likely that having
a stick shift car will save you any money.
